---
title: "Lab 4"
author: "Richard Romano"
output: pdf_document
date: "11:59PM Feb 29, 2020"
---

We now move on to simple linear modeling using the ordinary least squares algorithm.

Let's quickly recreate the sample data set from practice lecture 7:

```{r}
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
y = beta_0 + beta_1 * x + rnorm(n, mean = 0, sd = 0.33)
```

Rewrite the computation of y so that it is h*(x) + epsilon.

```{r}
h_star_x = beta_0 + beta_1 * x 
epsilon = rnorm(n, mean = 0, sd = 0.33)
y = h_star_x + epsilon
```

Graph the data by running the following chunk:

```{r}
pacman::p_load(ggplot2)
simple_df = data.frame(x = x, y = y)
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
simple_viz_obj
```

Does this make sense given the values of $beta_0$ and $beta_1$?

Yes.

Write a function `my_simple_ols` that takes in a vector `x` and vector `y` and returns a list that contains the `b_0` (intercept), `b_1` (slope), `yhat` (the predictions), `e` (the residuals), `SSE`, `SST`, `MSE`, `RMSE` and `Rsq` (for the R-squared metric). Internally, you can only use the functions `sum` and `length` and other basic arithmetic operations. You should throw errors if the inputs are non-numeric and not the same length. You should also name the class of the return value 'my_simple_ols_obj` by using the `class` function as a setter. No need to create ROxygen documentation here.

```{r}
my_simple_ols = function(x,y){
 if (class(x) != "numeric" | class(y) != "numeric") {stop("argument x or y is not numeric")}
 n = length(x)
  if (n != length(y)){stop("x and y must be same length")}
 
  y_bar = sum(y)/length(y)
  x_bar = sum(x)/length(x)
  s_x_squared = (1/(n-1) * sum((x - x_bar)^2))
  s_xy = (1/(n-1)) * sum((x - x_bar)*(y - y_bar))
  b1= s_xy/s_x_squared
  b0= y_bar - b1*x_bar
  y_hat = b0 + b1*x
  
  e = y - y_hat
  SSE = sum(e^2)
  SST = sum((y-y_bar)^2)
  Rsq = 1 - SSE/SST # 1 - NA
  MSE = SSE / (n-2)
  RMSE = sqrt(MSE)
  
  mod = list(b_0 = b0, b_1 = b1, y_hat = y_hat, e = e, SSE = SSE, SST = SST, Rsq = Rsq, MSE = MSE, RMSE = RMSE)
  class(mod) = "my_simple_ols_obj"
  mod
}
```

Verify your computations are correct for the vectors `x` and `y` from the first chunk using the `lm` function in R:

```{r}
lm_mod = lm(y ~ x)
my_lm_mod = my_simple_ols(x, y)
#run the tests to ensure the function is up to spec
pacman::p_load(testthat)
expect_equal(my_lm_mod$b_0, as.numeric(coef(lm_mod)[1]), tol = 1e-4)
expect_equal(my_lm_mod$b_1, as.numeric(coef(lm_mod)[2]), tol = 1e-4)
expect_equal(my_lm_mod$RMSE, summary(lm_mod)$sigma, tol = 1e-4)
expect_equal(my_lm_mod$Rsq, summary(lm_mod)$r.squared, tol = 1e-4)
```

Verify that the average of the residuals is 0.

```{r}
expect_equal(mean(my_lm_mod$e), 0, tol = 1e-4)
```

Create the $X$ matrix for this data example.

```{r}
X = cbind(1, x)
X
```

Use the `model.matrix` function to compute the matrix `X` and verify it is the same as your manual construction.

```{r}
model.matrix(~ x)
```

Using matrix algebra, verify the OLS estimate is the same as you computed from the `my_simple_ols` function. 

```{r}
XtXinvX = solve(t(X) %*% X) %*% t(X)
b = XtXinvX %*% y
b
```

Find the hat matrix $H$.

```{r}
H = X %*% XtXinvX
H
```

Verify that this specific hat matrix is symmetric.

```{r}
expect_equal(H, t(H))

```


Using the `diag` function, find the trace of the hat matrix.

```{r}
diag(H)
sum(diag(H))
```


Create a prediction method `g` that takes in a vector `x_future` and `my_simple_ols_obj`, an object of type `my_simple_ols_obj` and predicts y values for each entry in `x_future`. 

```{r}
g = function(x_future, my_simple_ols_obj){
  my_simple_ols_obj$b_0 + my_simple_ols_obj$b_1 * x_future
}
```

Use this function to verify that when predicting for the average x, you get the average y.

```{r}
expect_equal(g(mean(x), my_lm_mod), mean(y))
```

Create a prediction method `g` that takes in a vector `x_future` and the dataset $\mathbb{D}$ i.e. `X` where the first column is the one vector and `y` and returns the OLS predictions.

```{r}
g = function(x_future, X, y){
b = solve(t(X) %*% X) %*% t(X) %*% y
b[1] +b[2]*x_future
}
```

In class we spoke about error due to ignorance, misspecification error and estimation error. Show that as n grows, estimation error shrinks. Let us define an error metric that is the difference between $b_0$ and $b_1$ and $\beta_0$ and $\beta_1$. How about $h = ||b - \beta||^2$ where the quantities are now the vectors of size two. Show as n increases, this shrinks. 

```{r}
ns = 10^(1:7)
errors = array(dim=length(ns))
beta = c(beta_0, beta_1)
for (i in 1:length(ns)) {
  n = ns[i]
  x = runif(n)
  h_star_x = beta_0 + beta_1 * x 
  epsilon = rnorm(n, mean = 0, sd = 0.33)
  y = h_star_x + epsilon
  
  mod = lm(y ~ x)
  b = coef(mod)
  errors[i] = sum((beta - b)^2)
}
errors
```


We are now going to repeat one of the first linear model building exercises in history --- that of Sir Francis Galton in 1886. First load up package `HistData`.

```{r}
pacman::p_load(HistData)
```

In it, there is a dataset called `Galton`. Load it up.

```{r}
data(Galton)
```

You now should have a data frame in your workspace called `Galton`. Summarize this data frame and write a few sentences about what you see. Make sure you report $n$, $p$ and a bit about what the columns represent and how the data was measured. See the help file `?Galton`.
#p is 1 and n is 928 the number of observations  

```{r}
Galton
summary(Galton)
length(Galton$parent)
length(Galton$child)
# The galton data frame is a famous regression model used to explain how tall parents have shorter kids and shor parents have taller kids, that is wheter the parents are tall or small, they regress (bad wording here) towards the mean. The term regression model was born.The columns represnet height for parents and height for children respectively. The data was recorded in class intervals with width 1.0, All of the heights of female children were multiplied by 1.08 before tabulation to compensate for sex differences. 
?Galton
```


Find the average height (include both parents and children in this computation).

```{r}
avg_height = (mean(Galton$parent) + mean(Galton$child)) / 2
```

If you were to use the null model, what would the RMSE be of this model be?

```{r}
rmse_null = sqrt(mean((Galton$child - avg_height) ^ 2))
```

Note that in Math 241 you learned that the sample average is an estimate of the "mean", the population expected value of height. We will call the average the "mean" going forward since it is probably correct to the nearest tenth of an inch with this amount of data.

Run a linear model attempting to explain the childrens' height using the parents' height. Use `lm` and use the R formula notation. Compute and report $b_0$, $b_1$, RMSE and $R^2$. Use the correct units to report these quantities.

```{r}
mod = lm(child ~ parent, Galton)
b_0 = mod$coefficients[1]
b_1 = mod$coefficients[2]
b_0
b_1
summary(mod)$sigma
summary(mod)$r.sq

```

Interpret all four quantities: $b_0$, $b_1$, RMSE and $R^2$.

#b_0 is the intercept and can sometimes be interepeted as the value you would predict for y (childs height) if x1 is 0. That would not be a logical choice here. The slope does not have much meaning in this case. 
#b_1 is interepeted by parents height. given one unit change in a parents height, the childs average height will increase by 0.6462906 units.
#RMSE is 2.238457, therefore 95 percent of childrens height can be identified within 4.476914 plus or minus inches
#R^2 is 0.21 meaning the proportion of variance explained is 21%. R^2 is generally a vague interpetation to begin with.

How good is this model? How well does it predict? Discuss.

#Going by R^2? The model is weak. But I tend to believe R^2 is a poor measure to decide on how "good" a model is. The RMSE says we can identify 95% of childrens height within plus or minus 4.4769 inches. This seems pretty useful and by looking at the data, correct. I would like to see additional information to see how good the model is, that is a scatter plot, possibly the regression line itself. But its not a terrible model, regardless of what R^2 says. 

It is reasonable to assume that parents and their children have the same height? Explain why this is reasonable using basic biology and common sense.

#yes this is a reasonable conjecture depending on what you mean by same. science tells us that genetics defintley plays a role in a persons height. Now a common sense approach, tells us that you would be hard pressed to find many people who both have tall parents end up being on the small and you would be hard pressed to find two small parents produce a tall chuld. It does happen but its unlikely. A common estimate that is used is, add the mothers and fathers height, add 5 more inches if the child is a boy, subtract 5 if the child is a girl and then divide by 2. This is an intresting estimation that has been created to estimate a persons height based off there parents.  

If they were to have the same height and any differences were just random noise with expectation 0, what would the values of $\beta_0$ and $\beta_1$ be?

#b_0 the intercept would be the same. b_1 would be equal to 1. Parents height would not change the childrens height

Let's plot (a) the data in $\mathbb{D}$ as black dots, (b) your least squares line defined by $b_0$ and $b_1$ in blue, (c) the theoretical line $\beta_0$ and $\beta_1$ if the parent-child height equality held in red and (d) the mean height in green.

```{r}
pacman::p_load(ggplot2)
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue", size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", size = 1) +
  geom_abline(intercept = avg_height, slope = 0, color = "darkgreen", size = 1) +
  xlim(63.5, 72.5) + 
  ylim(63.5, 72.5) +
  coord_equal(ratio = 1)
```

Fill in the following sentence: 

TO-DO: Children of short parents became taller then the parents on average and children of tall parents became shorter then the parents on average.

Why did Galton call it "Regression towards mediocrity in hereditary stature" which was later shortened to "regression to the mean"?
Galton observed that extreme charteristics in parents are not passed on completket to their offspring. The characterstics in the offspring regress towards a medicore point. That is the mean here. 

Why should this effect be real?
#This makes sense because tall parents can not be more likely to have children taller then them, they are considered "tall" for a reason, that is it is a defined area on the extreme part of the data being observed. With new data coming in, the children, it makes sense for them to be more likely shorter then the extreme case and more towards the average case, that is it is more likely, higher probability to be shorter then tall parents. The same argument can be said for short parents and how children are likely to be taller.

You now have unlocked the mystery. Why is it that when modeling with $y$ continuous, everyone calls it "regression"? Write a better, more descriptive and appropriate name for building predictive models with $y$ continuous.

#Continious linear model. Not a fancy name, but to the point and accurate. 

Create a dataset $\mathbb{D}$ which we call `Xy` such that the linear model as $R^2$ about 50\% and RMSE approximately 1.

```{r}
x = c(2,3,4,5)
y = c(4,11,40,25)
Xy = data.frame(x = x, y = y)
Xy
E = lm(x ~ y, Xy)
summary(E)$sigma
summary(E)$r.sq

```

Create a dataset $\mathbb{D}$ which we call `Xy` such that the linear model as $R^2$ about 0\% but x, y are clearly associated.

```{r}
x = c(49.667,49.667,49.667,49.667,49.667,49.667,49.6667,49.667,49.667,49.667,49.667,49.667)
y =c(2, 10, 25, 39, 54, 29,48, 79,100,5,85,120)
Xy = data.frame(x = x, y = y)
L = lm(x ~ y, Xy)
summary(L)$r.sq

```

Extra credit: create a dataset $\mathbb{D}$ and a model (hint: not a lienar model) that can give you $R^2$ arbitrarily close to 1 but RMSE arbitrarily high.

```{r}
x = c(exp(3),exp(5),exp(7),exp(9),exp(10), exp(12), exp(15))
y = c(gamma(3),gamma(5),gamma(7),gamma(9),gamma(10), gamma(12), gamma(15))
Xy = data.frame(x = x, y = y)
P = lm(x ~ y, Xy)
summary(P)$sigma
summary(P)$r.sq

```

Load up the famous iris dataset. We are going to do a different prediction problem. Imagine the only input x is Species and you are trying to predict y which is Petal.Length. What would a reasonable, naive prediction be under all Species? Hint: it's what we did in class.

```{r}
data(iris)
iris


```

Prove that this is the OLS model by fitting an appropriate `lm` and then using the predict function to verify you get the same answers as you wrote previously. Show this by doing a linear regression with and without the intercept.

```{r}
iris$Species = factor(iris$Species)
is.factor(iris$Species)
mod = lm(Petal.Length ~ Species, iris)
mod_no_intercept = lm(Petal.Length ~ 0 +Species, iris)
summary(mod)
summary(mod_no_intercept)
predict(mod, Species = seretosa)
```

Use the `model.matrix` function to compute the matrix `X` for the regression with the intercept and without the intercept. What is different?

```{r}
X = model.matrix(mod, iris)
X
X1 = model.matrix(mod_no_intercept, iris)
X1
#sertosa is the intercept in the model with the intercept whereas all 3 species are in the no intercept model
```







